---
params:
  subtitle: ""
  title: ""
  authors: ""
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r lubridate::now()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # do not echo code
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(message = TRUE)
knitr::opts_chunk$set(fig_caption = TRUE)
knitr::opts_chunk$set(fig_height = 6) # default, make it bigger to stretch vertical axis
knitr::opts_chunk$set(fig_width = 8) # full width
knitr::opts_chunk$set(tidy = TRUE) # tidy up code in case echo = TRUE
```

```{r codeSetup, include=FALSE}

# library(SAVEr) # already loaded in makeFile

# Set start time ----
startTime <- proc.time()

# What are we running on?
sysInfo <- Sys.info()
sysName <- sysInfo[[1]]
nodeName <- sysInfo[[4]]
userName <- sysInfo[[7]]

# Functions ----

# Where are we?

# give feedback
print(paste0("Running on ", sysName, " with projLoc = ", myParams$projLoc))

# Local libraries ----
library(ggplot2) # for fancy graphs
library(kableExtra) # for better kable
library(lubridate) # for date stuff
library(skimr) # for data dictionaries

# Local parameters ----

# http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
# with grey
myParams$cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# with black
myParams$cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

\newpage

> Citation

If you wish to use any of the material from this report please cite as:

 * Anderson, B.(`r lubridate::year(lubridate::now())`) `r params$title`: `r params$subtitle`, SERG: University of Southampton.

\newpage

# About

Circulation:

 * Public

## Purpose

This report is intended to: 

 * estimate the number of off-grid households in Scotland;
 * with a specific focus on the Highlands & Islands

This work was carried out by the [Sustainable Energy Research Group](http://www.energy.soton.ac.uk) in the University of Southampton's [Faculty of Engineering and Environment](http://www.southampton.ac.uk/engineering/index.page). 

## History

Generally tracked via [github](https://github.com/dataknut/offGrid/commits/master)
 
## Re-use

The work is (c) `r lubridate::year(lubridate::now())` the University of Southampton.

The work is published under the Creative Commons Attribution-ShareAlike 4.0 International ([CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/)) License.

 * You are free to:
    * Share — copy and redistribute the material in any medium or format;
    * Adapt — remix, transform, and build upon the material;
 * for any purpose, even commercially.


Under the following terms:

 * Attribution — You must give appropriate credit, provide a link to the license, and indicate if changes were made. You may do so in any reasonable manner, but not in any way that suggests the licensor endorses you or your use. 
 * ShareAlike — If you remix, transform, or build upon the material, you must distribute your contributions under the same license as the original.
 * No additional restrictions — You may not apply legal terms or technological measures that legally restrict others from doing anything the license permits.

Re-use of the software code contained within this work is governed by an additional [license](https://github.com/dataknut/offGrid/blob/master/LICENSE).

\newpage

# Introduction

It started with a tweet:

 * https://twitter.com/hannon_matthew/status/1242790919818153984

# Data used

 * Census 2011 data on the [number of household spaces for Scotland at datazone level](https://www.scotlandscensus.gov.uk/ods-web/download/getDownloadFile.html?downloadFileIds=SNS%20Data%20Zone%202011%20blk);
 * BEIS 2010 - 2018 data on the [number of electricty meters at datazone level](https://www.gov.uk/government/collections/sub-national-electricity-consumption-data).

# Method 1: counting meters

If we assume:

 * the number of household spaces off grid = n household spaces - n meters

Then in theory we can estimate the number who are off grid. Of course some may be 'off grid' but still have a meter. And it is crucial that we have an accurate estimate of the number of household spaces to match to the BEIS data. Of course we only have that at datazone level for 2011...as we will see.

Here's the code - just to be sure!

```{r calc, echo=TRUE}
# subtract n elec meters from n household spaces

t <- censusDT[, .(zonecode, `Unoccupied household spaces: Total`, `Occupied household spaces`, `All household spaces`, nHHspaces2011)]
t <- t[order(zonecode)]

kableExtra::kable(head(t), caption = "Census data check (first 6 rows)") %>%
  kable_styling()

t <- beisDT[, .(zonecode, METERS)]
t <- t[order(zonecode)]

kableExtra::kable(head(t), caption = "BEIS data check (first 6 rows)") %>%
  kable_styling()

scotlandDT[, estimatednHHSpacesOffGrid := nHHspaces2011 - METERS]
scotlandDT[, estimatedPcHHSpacesOffGrid := 100 * ( estimatednHHSpacesOffGrid/nHHspaces2011)]
```


## Results

### 2011

This should be the 'most accurate' year as the data sources match.

```{r test2011, fig.cap="Histogram of Estimated number of off grid household spaces in Scotland in 2011"}

dt2011 <- scotlandDT[year == 2011]

ggplot2::ggplot(dt2011, aes(x = estimatednHHSpacesOffGrid)) +
  geom_histogram() +
  labs(x = "Estimated number of off grid household spaces")

```

Well, that's already interesting. We seem to have more meters than household spaces in a lot of places.

```{r test2011pc, fig.cap="Histogram of estimated percentage of household spaces which are off grid in Scotland in 2011"}

ggplot2::ggplot(dt2011, aes(x = estimatedPcHHSpacesOffGrid)) +
  geom_histogram() +
  labs(x = "Estimated % of household spaces which are off-grid")

```

Naturally this turns itself into some quite large negative percentages (Figure \@ref(fig:test2011pc)). Where are they and do they make any sense?

```{r negTable2011}
t <- dt2011[,.(LAName,LSOAName, METERS, nHHspaces2011,
                    estimatedPcHHSpacesOffGrid)]

t <- t[order(estimatedPcHHSpacesOffGrid)]

kableExtra::kable(head(t, 10), caption = "Ten largest negative estimates") %>%
  kable_styling()
```

```{r negSummary2011, fig.cap = "Boxplot of estimated % of households off-grid by local authority"}
t <- dt2011[,.(nZones = .N, meanMeters = mean(METERS),
               meanHHspaces = mean(nHHspaces2011),
               mean_pc_offgrid = mean(estimatedPcHHSpacesOffGrid),
               min_pc_offgrid = min(estimatedPcHHSpacesOffGrid),
               max_pc_offgrid = max(estimatedPcHHSpacesOffGrid)),
            keyby = .(LAName)]

t <- t[order(mean_pc_offgrid)]

kableExtra::kable(head(t, 10), caption = "Summary by local authority") %>%
  kable_styling()

ggplot2::ggplot(dt2011, aes(x = LAName,y = estimatedPcHHSpacesOffGrid)) +
  geom_boxplot() +
  coord_flip()
```

Well, Figure \@ref(fig:negSummary2011) makes no sense to me at all. It suggets that the median off-grid % in all local authorities in Scotland is slightly negative. Shetland & Orkneys are outliers. We might have expected them to be outliers in the other direction. Did we get our maths wrong? Not according to Table \@ref(tab:negTable2011)...

### 2018

Let's got to the other extreme and repeat. By now we are 7 years since Census 2011 so our household space counts should be well out of date.


```{r test2018, fig.cap="Histogram of Estimated number of off grid household spaces in Scotland in 2018"}

dt2018 <- scotlandDT[year == 2018]

ggplot2::ggplot(dt2018, aes(x = estimatednHHSpacesOffGrid)) +
  geom_histogram() +
  labs(x = "Estimated number of off grid household spaces")

```

Much the same.

```{r test2018pc, fig.cap="Histogram of estimated percentage of household spaces which are off grid in Scotland in 2018"}

ggplot2::ggplot(dt2018, aes(x = estimatedPcHHSpacesOffGrid)) +
  geom_histogram() +
  labs(x = "Estimated % of household spaces which are off-grid")

```

Figure \@ref(fig:test2018pc) has got some very large negative values!

```{r negTable2018}
t <- dt2018[,.(LAName, LSOAName, METERS, nHHspaces2011,
                    estimatedPcHHSpacesOffGrid)]

t <- t[order(estimatedPcHHSpacesOffGrid)]

kableExtra::kable(head(t, 10), caption = "Ten largest negative estimates") %>%
  kable_styling()
```

```{r negSummary2018, fig.cap = "Boxplot of estimated % of households off-grid by local authority"}
t <- dt2018[,.(nZones = .N, meanMeters = mean(METERS),
               meanHHspaces = mean(nHHspaces2011),
               mean_pc_offgrid = mean(estimatedPcHHSpacesOffGrid),
               min_pc_offgrid = min(estimatedPcHHSpacesOffGrid),
               max_pc_offgrid = max(estimatedPcHHSpacesOffGrid)),
            keyby = .(LAName)]

t <- t[order(mean_pc_offgrid)]

kableExtra::kable(head(t, 10), caption = "Summary by local authority") %>%
  kable_styling()

ggplot2::ggplot(dt2011, aes(x = LAName,y = estimatedPcHHSpacesOffGrid)) +
  geom_boxplot() +
  coord_flip()
```

Figure \@ref(fig:negSummary2018) is just as curious...

Did we get our maths wrong? Not according to Table \@ref(tab:negTable2018)...

## Conclusion

Don't you hate it when a good idea goes belly-up.

So what's going on? The [BEIS methodology report](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/853104/sub-national-methodology-guidance.pdf) says:

> The number of meters does not exactly equal the number of properties. The reasons for this are as follows:

> 1. An apartment building may have a meter for the building complex (used to power communal electricity uses such as lighting) in addition to each individual apartment having its own meter.
  
> 2. Some households may have a 3-rate meter system. A household with such a system will have one meter which measures all consumption at a peak rate and another meter which measures two other rates of off-peak consumption. This is the case for many households in Scotland, but it is less common to find a similar case in England or Wales.

So that's why...

Now what?

 * ask BEIS to provide a count of properties with meters instead of just a count of meters? This must be possible because the [BEIS methodology document](https://www.gov.uk/government/publications/regional-energy-data-guidance-note) states:
 
> "Meter postcode address file from ECOES: Geographic data is obtained for each meter from the Electricity Central Online Enquiry Service (ECOES) data file, which includes the address and postcode of each MPAN."

# Method 2 - using 'old' data

Previous versions of the LSOA/datazone level data (which no longer seem to be available from the BEIS [website](https://www.gov.uk/government/statistics/lower-and-middle-super-output-areas-electricity-consumption)) had more detail:

```{r checkOld2011}
names(oldElec2011DT)

oldElec2011DT[, zonecode := `Lower Layer Super Output Area (LLSOA) Code`]
setkey(oldElec2011DT, zonecode)

head(oldElec2011DT)

dt <- scotlandDT[year == 2011, .(METERS, zonecode, LAName)]
setkey(dt, zonecode)

message("New data: zonecodes beginning with 'S' (Scotland)")
head(dt[zonecode %like% "S"])

message("Old data: zonecodes beginning with 'S' (Scotland)")
head(oldElec2011DT[zonecode %like% "S"])

```

Nope. The old data does not include Scotland...

# Annex

## Data summaries

Census

```{r skim.census}
skimr::skim(censusDT)
```

BEIS

```{r skim.beis}
skimr::skim(beisDT)
```


# Runtime


```{r check runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes) using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform`.

# R environment

R packages used:

 * base R - for the basics [@baseR]
 * data.table - dataframes on steroids [@data.table]
 * ggplot2 - for slick graphics [@ggplot2]
 * kableExtra - fancy tables [@kableExtra]
 * knitr - to create this document [@knitr]
 * lubridate - date stuff [@lubridate]
 * skimr - data summaries [@skimr]
 
```{r sessionInfo}
sessionInfo()
```

# References

